---
title: "Research Project 2 - STH in Ethiopia"
output: github_document

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(sf)
library(raster)
library(tmap)
library(lme4)
library(PrevMap)
library(scico)

sth <- read.csv("data/ETH_sth.csv")
sth$Asc_examined <- as.numeric(sth$Asc_examined)

sth_sf <- st_as_sf(sth, coords = c("utm_x", "utm_y"))
st_crs(sth_sf) <- 32638

ETH_grid <- read.csv("data/ETH_grid.csv")


ETH_adm0 <- st_read("ETH_files/ETH_adm/ETH_adm0.shp")
ETH_adm0 <- st_transform(ETH_adm0, crs = 32638)

ETH_adm1 <- st_read("ETH_files/ETH_adm/ETH_adm1.shp")
ETH_adm1 <- st_transform(ETH_adm1, crs = 32638)

ETH_adm2 <- st_read("ETH_files/ETH_adm/ETH_adm2.shp")
ETH_adm2 <- st_transform(ETH_adm2, crs = 32638)



```

# Data sources 

- STH data: [ESPEN](https://espen.afro.who.int/tools-resources/download-data)
- Altitude and landcover: [DIVA GIS](https://www.diva-gis.org/)
- Friction surface and travel times: [Malaria Atlas Project](https://malariaatlas.org/)
  - Friction surface recorded as minutes required to travel one meter, walking only and with motorized vehicle
  - Travel time to nearest healthcare centre, in minutes 

```{r STH_points, echo=TRUE }

ggplot()+
  geom_sf(data = ETH_adm1, fill = NA, col = "grey")+
  geom_sf(data = ETH_adm0, fill = NA, col = "black")+
  geom_sf(data = sth_sf, size = 0.5, col = "red")+
  coord_sf()+
  theme_void()

```


# STH Prevalence 

WHO classifies prevalence into [three risk groups](https://apps.who.int/iris/bitstream/handle/10665/44804/9789241503129_eng.pdf), requiring mass-drug administrations at different frequencies. 
These risk groups are classified as follows:

- High risk, >= 50% prevalence of any STH. Requires treatment twice a year.
- Moderate risk, < 50% and >= 20% prevalence of any STH. Requires treatment once a year. 
- Low risk, < 20% prevalence of any STH. No mass treatment required, done on a case-by-case basis. 

# Models 

## Generalised linear mixed-effects model (binomial)

$$ \log\left(\frac{p}{1-p}\right) = x_i^\intercal\beta + Z_i$$

## Geostatistical model (binomial)

$$ \log\left(\frac{p}{1-p}\right) = x_i^\intercal\beta + S(x_i) + Z_i $$

# Predicted prevalence 

## Hookworms 

